---
title: "Compare the Proportion of infected Mosquitoes Over Time"
output: pdf_document
date: ""
---

```{r echo = F, warning=FALSE, message=FALSE}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```


## 1. Chi-square Test for Aedes albopictus

We used a Chi-square test to see if the proportion of mosquitoes testing positive for infection, dissemination, or transmission changes across dpi (days post-infection) for Aedes albopictus. The test compares observed counts to what we’d expect if dpi had no effect. The null hypothesis says the proportion of positive mosquitoes stays the same over time. We found a significant difference for infection ($p < 0.001$), but not for dissemination ($p = 0.066$) or transmission ($p = 0.111$). This suggests infection changes with time, but no significant change for dissemination and transmission.

```{r echo = F, warning=FALSE, message=FALSE}
# Load data
# Load data
df <- readxl::read_excel("madv_df_03_27_2025_for Martial.xlsx")

# Prepare indicators
df <- df %>%
  mutate(
    Infection = ifelse(Tissue == "Body" & qRT_PCR_result == "Pos", 1, 0),
    Dissemination = ifelse(Tissue == "Legs" & qRT_PCR_result == "Pos", 1, 0),
    Transmission = ifelse(Tissue == "Saliva" & qRT_PCR_result == "Pos", 1, 0),
    dpi = as.numeric(dpi)
  )

# Filter for Aedes albopictus
albopictus_df <- df %>%
  filter(Mosquito_specie == "Aedes_albopictus")

# Chi-square test function
run_chi2 <- function(data, tissue, outcome) {
  tbl <- table(data %>% filter(Tissue == tissue) %>% select(dpi, all_of(outcome)))
  result <- chisq.test(tbl)
  tibble::tibble(
    Category = outcome,
    Statistic = round(result$statistic, 2),
    df = result$parameter,
    p_value = signif(result$p.value, 3)
  )
}

# Run tests and combine results
results_table <- dplyr::bind_rows(
  run_chi2(albopictus_df, "Body", "Infection"),
  run_chi2(albopictus_df, "Legs", "Dissemination"),
  run_chi2(albopictus_df, "Saliva", "Transmission")
)

# Print ltable
results_table %>%
  kable(format = "latex", booktabs = TRUE, caption = "Chi-square test results for Aedes albopictus") %>%
  kable_styling(latex_options = c("hold_position", "striped"))

```

```{r echo = F, warning=FALSE, message=FALSE, fig.height=3, fig.width=6}
# Prepare proportions by dpi
plot_df <- albopictus_df %>%
  mutate(Category = case_when(
    Tissue == "Body" ~ "Infection",
    Tissue == "Legs" ~ "Dissemination",
    Tissue == "Saliva" ~ "Transmission"
  )) %>%
  group_by(
    Category, dpi
    ) %>%
  summarise(
    Proportion_Positive = mean(qRT_PCR_result == "Pos"),
    .groups = "drop"
    )

ggplot(plot_df, aes(x = dpi, y = Proportion_Positive, color = Category)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Aedes albopictus: Infection, Dissemination, and Transmission Rates",
    x = "DPI",
    y = "Proportion Positive",
    color = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

```


