---
title: "Statistical Test for Detecting Differences Between Two Time Trends"
output: pdf_document
date: ""
---

```{r echo = F, warning=FALSE, message=FALSE}
# Load libraries
library(dplyr)
library(ggplot2)
library(lme4)
library(readxl)
```
## 1. Exploratory Overview of the Experiment

This dataset originates from a vector competence experiment and investigating whether two mosquito species: 1. Aedes aegypti and 2. Aedes albopictus can become infected with and transmit a virus over time.

```{r echo = F, warning=FALSE, message=FALSE}
# Load and prepare data
df <- read_excel("madv_df_03_27_2025_for Martial.xlsx")

df_plot <- df %>%
  mutate(
    Infected = ifelse(qRT_PCR_result == "Pos", 1, 0),
    Mosquito_specie = as.factor(Mosquito_specie),
    dpi = as.numeric(dpi)
  )

# Aggregate infection rate by species and dpi
trend_df <- df_plot %>%
  group_by(Mosquito_specie, dpi) %>%
  summarise(
    Infection_Rate = mean(Infected)
  )

# Plot
ggplot(trend_df, aes(x = dpi, y = Infection_Rate, color = Mosquito_specie)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Infection Rate Over Time by Mosquito Species",
    x = "Days Post-Infection (DPI)",
    y = "Proportion Infected",
    color = "Mosquito Species"
  ) +
  theme_minimal()
```

### Experimental Design Summary

- *Mosquito species*: Aedes aegypti, Aedes albopictus.
- *Time points*: 3, 7, 14, and 21 days post-infection (dpi)
- *Replicates*: 2 biological replicates
- **Tissue types sampled**:
  - `Body`: assesses infection
  - `Legs`: assesses dissemination
  - `Saliva`: assesses transmission
- **Virus detection method**: qRT-PCR
  - Binary outcome: `Pos` (positive) or `Neg` (negative)

### Variables of Interest

- **Predictors**: mosquito species (`Mosquito_specie`), days post-infection (`dpi`)
- **Outcomes**:
  - **Infection** – determined by qRT-PCR result from `Body`
  - **Dissemination** – determined by qRT-PCR result from `Legs`
  - **Transmission** – determined by qRT-PCR result from `Saliva`

### Main Research Question

> Do the temporal trends of infection, dissemination, and transmission significantly differ between Aedes aegypti and Aedes albopictus?

---

## 2. Statistical Modeling

### 2.1 Logistic Regression for Infection

To evaluate the effect of mosquito species and dpi on infection status, we fit a logistic regression model.

### Hypotheses

- **Null hypothesis** ($H_0$): The effect of dpi on infection probability is the same for both species.
- **Alternative hypothesis** ($H_1$): The effect of dpi on infection probability differs between species.


### Model Formula

We model the log-odds of infection using the logistic regression model:

$$
\log\left(\frac{p}{1 - p}\right) = \beta_0 + \beta_1 \cdot \text{Species} + \beta_2 \cdot \text{dpi} + \beta_3 \cdot (\text{Species} \times \text{dpi})
$$

 - where $p$ is the probability of infection (i.e., positive qRT-PCR result for body tissue).




```{r echo = F, fig.height=4, include=FALSE}
# Filter and prepare data
df <- readxl::read_excel("madv_df_03_27_2025_for Martial.xlsx")
df_cleaned <- df %>%
  mutate(
    Infection = ifelse(Tissue == "Body" & qRT_PCR_result == "Pos", 1, 0),
    Mosquito_specie = as.factor(Mosquito_specie),
    dpi = as.numeric(dpi)
  ) %>%
  filter(Tissue == "Body")

# Fit logistic regression
model <- glm(Infection ~ Mosquito_specie * dpi,
             family = binomial, 
             data = df_cleaned)

# Predicted probabilities
df_cleaned$predicted_prob <- predict(model, type = "response")


# Plot predicted probabilities
ggplot(df_cleaned, 
       aes(x = dpi, y = predicted_prob, color = Mosquito_specie)
       ) +
  geom_line(
    stat = "summary",
    fun = mean
    ) +
  geom_point(
    alpha = 0.3
    ) +
  labs(
    title = "Predicted Infection Probability by DPI and Species",
    y = "Predicted Probability",
    x = "Days Post-Infection (DPI)") +
  theme_minimal()
# summary(model)

```






### 2.2 Mixed-Effects Logistic Regression for Infection

To consider for experimental replication, we extend the previous model by fitting a mixed-effects logistic regression model. This allows us to model infection probability while adjusting for potential variation between biological replicates.

Hypotheses:

- Null hypothesis ($H0$): The effect of dpi on infection probability is the same for both species, and replicates do not influence the results.
- Alternative hypothesis ($H_1$): The effect of dpi on infection probability differs between species, after adjusting for replicate-level variability.


###  2.3 Bayesian Logistic Regression for Infection

We fit a Bayesian logistic regression model. This approach allows estimating the probability that infection rates differ by species and time, using Bayesian inference.

Hypotheses:

- Null hypothesis ($H_0$): There is no effect of dpi, species, or their interaction on infection probability.
- Alternative hypothesis ($H_1$): There is an effect of at least one predictor on infection probability, with uncertainty expressed via posterior distributions.

